A Stopped Negative Binomial Distribution

### 2016-06-16 stat-and-prob-letters/snb-paper.tex
Release candidate?

%% or include affiliations in footnotes:
\author[mymainaddress]{Michelle DeVeaux}
\ead{michelle.deveaux@yale.edu}

\author[mymainaddress]{Michael J. Kane\corref{mycorrespondingauthor}}
\cortext[mycorrespondingauthor]{Corresponding author}
\ead{michael.kane@yale.edu}

\author[mymainaddress]{Daniel Zelterman}
\ead{daniel.zelterman@yale.edu}

\address[mymainaddress]{Department of Biostatistics\\ School of Epidemiology and Public Health\\ Yale University, New Haven, CT}

@@

\section{Introdution and Motivation}

@@

If all 12 patients are enrolled at once, as in the classic
design, then the sample size is 12. However, in most clinical trials the
patients are enrolled sequentially, often with one patient's outcome realized
before the next one enters the trial. In the present example, observing two
successful patients allows us to reach one endpoint so the sample required
could be as small as two. Similarly 11
observed treatment failures also ends the trial. This sampling mechanism, in
which the experiment ends as soon as a predefined endpoint is reached, is
call {\em curtailed sampling}. Under curtailed sampling the range of the
sample size for this trial is between two and 12.

@@

The probability a given realization of $\mathbf{X}$ reaches $s$ at
the $k$th outcome is the probability that, at time $k-1$, there are $s-1$
successful outcomes and $k-s$ unsuccessful outcomes multiplied by
the probability of the final success at time $k$. This expression is given
in (\ref{eqn:N}). 
Similarly, the probability a given realization reaches $k-t$
is the probability that, at outcome $k-1$ there are $k-t$ successful outcomes
and $t-1$ unsuccessful outcomes multiplied by the probability of a final
unsuccessful outcome at time $k$.

@@

Let $\mathcal{I}_p(s, t)$ be the {\em regularized incomplete beta function} 
\citep{Olver2010}. This function satisfies 
$\mathcal{I}_p(s, t) = 1-\mathcal{I}_{1-p}(t, s)$ \citep{Uppuluri1970}. Then
\begin{align*}
R = \sum_{i=0}^{t-1} &{i+s-1 \choose i} p^s (1-p)^i +
\sum_{j=0}^{s-1}  {j+t-1 \choose j} p^j  (1-p)^t \\
   &= 1-\mathcal{I}_p(s, t) + 1 - \mathcal{I}_{1-p}(t, s) \\
   &= 1. 
\end{align*}
This completes the proof (\ref{eqn:pmf}) is the distribution of the
stopping time and is a valid probability mass function.
\end{proof}

@@

The SNB also generalizes both the minimum (riff-shuffle) and maximum negative
binomial distributions up to a translation of the support.
For the special case of $\,s=t,$ the distribution of $\,Y\,$ is the
riff-shuffle, or minimum negative binomial distribution~\citep{Uppuluri1970}.
Similar derivations of the closely-related maximum negative binomial 
discrete distributions also appear in~\cite{Zhang2000}
and \cite{Zelterman2005}.
The maximum negative binomial is the smallest number of outcomes necessary to 
observe at least $s$ successes {\em and} $s$ failures. The SNB is the 
number of outcomes to observe {\em either} $s$ successes or $t$ failures.

@@

To illustrate this result, let us return to our initial example
where $s=2$, $t=12$, and $p=0.045$.  The probability masses in
Fig.~\ref{fig:snb_bin_compare} represented in 
black are equal in panels a) and b) as are the masses in grey.
The probability that $s$
successes are reached in the SNB process is the same as the binomial 
probability of at least two successes. Likewise, the probability that $t$ 
failures are reached in the SNB process is the same as the binomial
probability of zero or one successes.

\begin{figure}[t!]
\centering
\subfloat[The SNB distribution.]{\includegraphics[width=0.5\textwidth]{snb_density.pdf}}
\hfill
\subfloat[The Binomial distribution.]{\includegraphics[width=0.5\textwidth]{bin_density.pdf}}
\caption{
SNB(0.045, 2, 12) with mass contributed from 
$s$ successes (black) or $t$ failures (grey) along with Bin(0.045, 12) with
at least 2 successes (black) or fewer (grey).
}
\label{fig:snb_bin_compare}
\end{figure}

@@

The SNB's ability to approximate the geometric, normal, and gamma 
distributions follow from it generalizing the negative binomial. To 
recover the MGF of the negative binomial consider the case where
$t \rightarrow \infty$ in (\ref{eqn:mgf}). The regularized incomplete
beta function in the first term goes to one and zero in the second term
and we are left with the MGF of the negative binomial distribution. 
When $s=1$ this is the same as the geometric distribution. The negative
binomial can therefore be seen as a sum of i.i.d. geometric distributions.
For an appropriately large number of samples the central limit theorem
hold yielding a normal approximation.

Drawing a connection to the gamma distribution is more complicated.
However it is a well-studied problem in the literature (see 
\cite{Ord1968,Guenther1972,Best1974} for examples). Methods for these
approximation generally work by equating cumulants and then showing 
that errors are not too far off.

@@

To better understand the compound distribution, let us once again return to 
our example where $s=2$, $t=12$. However, we assume $p$ is a Beta distribution
with shape parameters $\alpha = 0.045 c$ and 
$\beta = 0.955 c$ for any $c > 0$.
This parameterization allows us to examine the relationship between 
uncertainty in $p$ and the SNB.

\begin{figure}[bp!]
\centering
\subfloat[The compound distribution's expected sample size.]{\includegraphics[width=0.5\textwidth]{bayesian-sample-expectation.pdf}}
\hfill
\subfloat[The compound distribution's variance.]{\includegraphics[width=0.5\textwidth]{bayesian-sample-variance.pdf}}
\caption{
The expected size and variance of the compound SNB with $s=2$, $t=12$
with $\mathbb{E}\ p = 0.045$ varying the uncertainty 
in the distribution of $p$. A larger value of $c$
corresponds to a Beta with smaller variance.
}
\label{fig:bayesian-sample-size}
\end{figure}

>>>>

%% or include affiliations in footnotes:
\author{Michelle DeVeaux}
\ead{michelle.deveaux@yale.edu}

\author{Michael J. Kane$^*$\footnote{* Corresponding author}}
\ead{michael.kane@yale.edu}

\author{Daniel Zelterman}
\ead{daniel.zelterman@yale.edu}

\address{Department of Biostatistics\\ School of Epidemiology and Public Health\\ Yale University, New Haven, CT}

@@

\section{Introduction and Motivation}

@@

If all 12 patients are enrolled at once, as in the classic
design, then the sample size is 12. However, in most clinical trials the
patients are enrolled sequentially, often with one patient's outcome realized
before the next one enters the trial. In the present example, observing two
successful patients allows us to reach one endpoint so the sample required
could be as small as two. Similarly 11
observed treatment failures also ends the trial. This sampling mechanism, in
which the experiment ends as soon as a predefined endpoint is reached, is
called {\em curtailed sampling}. Under curtailed sampling the range of the
sample size for this trial is between two and 12.

@@

The probability a given realization of $\mathbf{X}$ reaches $s$ at
the $k$th outcome is the probability that, at time $k-1$, there are $s-1$
successful outcomes and $k-s$ unsuccessful outcomes multiplied by
the probability of the final success at time $k$. This expression is given
in (\ref{eqn:N}). 
Similarly, the probability a given realization reaches $k-t$
is the probability that, at outcome $k-1$, there are $k-t$ successful outcomes
and $t-1$ unsuccessful outcomes multiplied by the probability of a final
unsuccessful outcome at time $k$.

@@

Let $\mathcal{I}_p(s, t)$ be the {\em regularized incomplete beta function} 
\citep{Olver2010}. This function satisfies 
$\mathcal{I}_p(s, t) = 1-\mathcal{I}_{1-p}(t, s)$ \citep{Uppuluri1970}. Then
\begin{align*}
R = \sum_{i=0}^{t-1} &{i+s-1 \choose i} p^s (1-p)^i +
\sum_{j=0}^{s-1}  {j+t-1 \choose j} p^j  (1-p)^t \\
   &= 1-\mathcal{I}_p(s, t) + 1 - \mathcal{I}_{1-p}(t, s) \\
   &= 1. 
\end{align*}
This completes the proof that (\ref{eqn:pmf}) is the distribution of the
stopping time and is a valid probability mass function.
\end{proof}

@@

The SNB also generalizes both the minimum (riff-shuffle) and maximum negative
binomial distributions up to a translation of the support.
For the special case of $\,s=t,$ the distribution of $\,Y\,$ is the
riff-shuffle, or minimum negative binomial distribution~\citep{Uppuluri1970}.
Similar derivations of the closely-related maximum negative binomial 
discrete distributions also appear in~\cite{Zhang2000}
and \cite{Zelterman2005}.
The maximum negative binomial is the smallest number of outcomes necessary to 
observe at least $s$ successes {\em and} $s$ failures. The SNB is the 
number of outcomes to observe {\em either} $s$ successes {\em or} $t$ failures.

@@

To illustrate this result, let us return to our initial example
where $s=2$, $t=11$, and $p=0.045$.  The probability masses in
Fig.~\ref{fig:snb_bin_compare} represented in 
black are equal in panels a) and b) as are the masses in grey.
The probability that $s$
successes are reached in the SNB process is the same as the binomial 
probability of at least two successes. Likewise, the probability that $t$ 
failures are reached in the SNB process is the same as the binomial
probability of zero or one successes.

\begin{figure}[t!]
\centering
\subfloat[The SNB distribution.]{\includegraphics[width=0.5\textwidth]{snb_density.pdf}}
\hfill
\subfloat[The Binomial distribution.]{\includegraphics[width=0.5\textwidth]{bin_density.pdf}}
\caption{
SNB(0.045, 2, 11) with mass contributed from 
$s$ successes (black) or $t$ failures (grey) along with Bin(0.045, 12) with
at least 2 successes (black) or fewer (grey).
}
\label{fig:snb_bin_compare}
\end{figure}

@@

The SNB's ability to approximate the geometric, normal, gamma, and poisson
distributions follow from it generalizing the negative binomial. To 
recover the MGF of the negative binomial consider the case where
$t \rightarrow \infty$ in (\ref{eqn:mgf}). The regularized incomplete
beta function in the first term goes to one and zero in the second term.
We are left with the MGF of the negative binomial distribution.

When $s=1$ this is the geometric distribution. The negative
binomial can therefore be seen as a sum of i.i.d. geometric distributions.
For an appropriately large number of samples the central limit theorem
yields a normal approximation.

Drawing connections to the gamma and poisson distributions are more complicated.
However a connection to the gamma distribution well-studied problem in the 
literature (see \cite{Ord1968,Guenther1972,Best1974} for examples). 
A connection to the poisson appears in \cite{Anscombe1950} where
it is shown that if the mean of a poisson is proportional
to a $\chi^2_{2k}$ distribution then the negative binomial is obtained. 
All of these results work by equating cumulants and then showing 
that differences between betwen the cumulant generating
functions converge to zero.

@@

To better understand the compound distribution, let us once again return to 
our example where $s=2$, $t=11$. However, we assume $p$ is a Beta distribution
with shape parameters $\alpha = 0.045 c$ and 
$\beta = 0.955 c$ for any $c > 0$.
This parameterization allows us to examine the relationship between 
uncertainty in $p$ and the SNB.

\begin{figure}[h!]
\centering
\subfloat[The compound distribution's expected sample size.]{\includegraphics[width=0.5\textwidth]{bayesian-sample-expectation.pdf}}
\hfill
\subfloat[The compound distribution's variance.]{\includegraphics[width=0.5\textwidth]{bayesian-sample-variance.pdf}}
\caption{
The expected size and variance of the compound SNB with $s=2$, $t=11$
with $\mathbb{E}\ p = 0.045$ varying the uncertainty 
in the distribution of $p$. A larger value of $c$
corresponds to a Beta with smaller variance.
}
\label{fig:bayesian-sample-size}
\end{figure}
